Integrated nested Laplace approximations for extended latent Gaussian models with application to the Naomi HIV model

### 2022-12-21 src/docs_paper/appendix.Rmd
Add naomi.cpp to appendix. Start writing main text model description
A few appendix fixes and extra into main text

#### HIV incidence rate

@@

For HIV prevalence and ART coverage we use binomial working likelihoods as follows
\begin{align}
Y^{\hat \rho}_{\{x\}, \{s\}, \{a\}, \nu} &\sim \text{xBin}(M^{\hat \rho}_{\{x\}, \{s\}, \{a\}, \nu}, \rho_{\{x\}, \{s\}, \{a\}, {T_1}}) \\
\rho_{\{x\}, \{s\}, \{a\}, t} &= \frac{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a} \cdot \rho_{x, s, a}}{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a}},
\end{align}
and
\begin{align}
Y^{\hat \alpha}_{\{x\}, \{s\}, \{a\}, \nu} &\sim \text{xBin}(M^{\hat \alpha}_{\{x\}, \{s\}, \{a\}, \nu}, \alpha_{\{x\}, \{s\}, \{a\}, {T_1}}) \\
\alpha_{\{x\}, \{s\}, \{a\}, t} &= \frac{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \alpha_{x, s, a}}{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a} \cdot \rho_{x, s, a}}.
\end{align}

@@

<!-- * Likelihood for number of ANC clients specified only^[Why?] for year $Y[T_2]$ as -->
<!-- $$ -->
<!-- W_{\{x\} Y[T_2]}^{\text{ANC}} \sim \text{Pois} \left( \frac{M_{Y[T 2]}^{\text{ANC}}}{12} \sum_{x \in\{x\}} \sum_{a \in\{15, \ldots 45\}} \Psi_{x, a, T_2} \right) -->
<!-- $$ -->

The observed number of HIV positive and already on ART among ANC clients at $Y[T_1]$ is modelled by
\begin{align*}
Y_{\{x\} Y[T_1]}^{\text{ANC}} &\sim \text{Bin} \left(X_{\{x\} Y[T_1]}^{\text{ANC}}, \rho_{\{x\},\{15, \ldots 45\}, t}^{\text{ANC}}\right) \\
Z_{\{x\} Y[T_1]}^{\text{ANC}} &\sim \text{Bin} \left(Y_{\{x\} Y[T_1]}^{\text{ANC}}, \alpha_{\{x\},\{15, \ldots 45\}, t}^{\text{ANC}}\right)
\end{align*}
where predicted prevalence and ART coverage are aggregated weighted by the predicted number of pregnant women by age $\Psi_{x, a}$
\begin{align*}
\rho_{\{x\}\{a\}, t}^{\text{ANC}} &= \frac{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a} \cdot \rho_{x, a}^{\text{ANC}}}{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a}} \\
\alpha_{\{x\}\{a\}, t}^{\text{ANC}} &= \frac{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a} \cdot \rho_{x, a}^{\text{ANC}} \cdot \alpha_{x, a}^{\text{ANC}}}{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a} \cdot \rho_{x, a}^{\text{ANC}}}
\end{align*}

@@

Let $\dot A_{\{x\}, \{s\}, \{a\}, t}$ be data for the number receiving ART
$$
\dot A_{\{x\}, \{s\}, \{a\}, t} = \sum_{s \in \{s\}} \sum_{a \in \{a\}} \sum_{x \in \{x\}} \sum_{x \sim x', x = x'} \dot A_{x', x, s, a},
$$
We model the unobserved numbers of ART clients travelling from $x'$ to $x$ as
$$
\dot A_{x', x, s, a} \sim \text{Bin}(N_{x', s, a}, \pi_{x', x, s, a})
$$
where $\pi_{x', x, s, a} = \rho_{x', s, a} \cdot \alpha_{x', s, a} \cdot \gamma_{x', x, s, a}$.
This likelihood is approximated using a normal for the sum of binomials by
$$
\dot A_{\{x\}, \{s\}, \{a\}, t} \sim \mathcal{N}(\tilde) A_{\{x\}, \{s\}, \{a\}, t}, \sigma^{\tilde A}_{\{x\}, \{s\}, \{a\}, t})
$$
where
$$
\tilde{A}_{\{x\},\{s\}\{a\}, t} = \sum_{s \in\{s\}} \sum_{a \in\{a\}} \sum_{x \in\{x\}} \sum_{x \sim x', x = x'} N_{x', s, a} \cdot \pi_{x', x, s, a},
$$
and
$$
\sigma_{\{x\},\{s\}\{a\}, t}^{\tilde{A}} = \sqrt{\sum_{s \in\{s\}} \sum_{a \in\{a\}} \sum_{x \in\{x\}} \sum_{x \sim x', x = x'} N_{x', s, a} \cdot \pi_{x', x, s, a} \cdot \left(1-\pi_{x', x, s, a}\right)}.
$$

@@

* If survey data on ART coverage by age and sex are not available then we set $u_a^\alpha = 0$ and $u_{a, s = \text{M}}^\alpha = 0$ and use the average age/sex pattern of ART coverage from the Spectrum offset $\eta_{R_x, s, a}^\alpha$.
* If no ART data (survey or ART programme) are available at $T_1$ but data on ART coverage among ANC clients are available, the level of ART coverage is not identifiable, but spatial variation is identifiable.
In this instance, overall ART coverage is determined by the Spectrum offset, and only area random effects are estimated
$$
\text{logit} \left(\alpha_{x, s, a, T_1} \right) = u_x^\alpha+\eta_{R_x, s, a}^\alpha \\
$$
* If survey data on recent HIV infection are not included in the model, then $\beta_0^\lambda = \beta_S^{\lambda, s = \text{M}} = u_x^\lambda = 0$. The sex ratio for HIV incidence is determined by the sex incidence rate ratio from Spectrum in the same years and the incidence rate in all districts is modelled assuming the same average HIV transmission rate for untreated adults, but varies according to district estimates of HIV prevalence and ART coverage.

>>>>

### HIV incidence rate

@@

For HIV prevalence and ART coverage we use binomial working likelihoods as follows
\begin{align}
Y^{\hat \rho}_{\{x\}, \{s\}, \{a\}, \nu} &\sim \text{xBin}(M^{\hat \rho}_{\{x\}, \{s\}, \{a\}, \nu}, \rho_{\{x\}, \{s\}, \{a\}, {T_1}}) \\
\rho_{\{x\}, \{s\}, \{a\}} &= \frac{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a} \cdot \rho_{x, s, a}}{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a}},
\end{align}
and
\begin{align}
Y^{\hat \alpha}_{\{x\}, \{s\}, \{a\}, \nu} &\sim \text{xBin}(M^{\hat \alpha}_{\{x\}, \{s\}, \{a\}, \nu}, \alpha_{\{x\}, \{s\}, \{a\}, {T_1}}) \\
\alpha_{\{x\}, \{s\}, \{a\}} &= \frac{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \alpha_{x, s, a}}{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a} \cdot \rho_{x, s, a}}.
\end{align}

@@

The observed number of HIV positive and already on ART among ANC clients at $Y[T_1]$ is modelled by
\begin{align*}
Y_{\{x\} Y[T_1]}^{\text{ANC}} &\sim \text{Bin} \left(X_{\{x\} Y[T_1]}^{\text{ANC}}, \rho_{\{x\},\{15, \ldots 45\}}^{\text{ANC}}\right) \\
Z_{\{x\} Y[T_1]}^{\text{ANC}} &\sim \text{Bin} \left(Y_{\{x\} Y[T_1]}^{\text{ANC}}, \alpha_{\{x\},\{15, \ldots 45\}}^{\text{ANC}}\right)
\end{align*}
where predicted prevalence and ART coverage are aggregated weighted by the predicted number of pregnant women by age $\Psi_{x, a}$
\begin{align*}
\rho_{\{x\}\{a\}}^{\text{ANC}} &= \frac{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a} \cdot \rho_{x, a}^{\text{ANC}}}{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a}} \\
\alpha_{\{x\}\{a\}}^{\text{ANC}} &= \frac{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a} \cdot \rho_{x, a}^{\text{ANC}} \cdot \alpha_{x, a}^{\text{ANC}}}{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a} \cdot \rho_{x, a}^{\text{ANC}}}
\end{align*}

@@

Let $\dot A_{\{x\}, \{s\}, \{a\}}$ be data for the number receiving ART
$$
\dot A_{\{x\}, \{s\}, \{a\}} = \sum_{s \in \{s\}} \sum_{a \in \{a\}} \sum_{x \in \{x\}} \sum_{x \sim x', x = x'} \dot A_{x', x, s, a},
$$
We model the unobserved numbers of ART clients travelling from $x'$ to $x$ as
$$
\dot A_{x', x, s, a} \sim \text{Bin}(N_{x', s, a}, \pi_{x', x, s, a})
$$
where $\pi_{x', x, s, a} = \rho_{x', s, a} \cdot \alpha_{x', s, a} \cdot \gamma_{x', x, s, a}$.
This likelihood is approximated using a normal for the sum of binomials by
$$
\dot A_{\{x\}, \{s\}, \{a\}} \sim \mathcal{N}(\tilde A_{\{x\}, \{s\}, \{a\}}, \sigma^{\tilde A}_{\{x\}, \{s\}, \{a\}})
$$
where
$$
\tilde{A}_{\{x\},\{s\}\{a\}} = \sum_{s \in\{s\}} \sum_{a \in\{a\}} \sum_{x \in\{x\}} \sum_{x \sim x', x = x'} N_{x', s, a} \cdot \pi_{x', x, s, a},
$$
and
$$
\sigma_{\{x\},\{s\}\{a\}}^{\tilde{A}} = \sqrt{\sum_{s \in\{s\}} \sum_{a \in\{a\}} \sum_{x \in\{x\}} \sum_{x \sim x', x = x'} N_{x', s, a} \cdot \pi_{x', x, s, a} \cdot \left(1-\pi_{x', x, s, a}\right)}.
$$

@@

* If survey data on ART coverage by age and sex are not available then we set $u_a^\alpha = 0$ and $u_{a, s = \text{M}}^\alpha = 0$ and use the average age/sex pattern of ART coverage from the Spectrum offset $\eta_{R_x, s, a}^\alpha$.
* If no ART data (survey or ART programme) are available at $T_1$ but data on ART coverage among ANC clients are available, the level of ART coverage is not identifiable, but spatial variation is identifiable.
In this instance, overall ART coverage is determined by the Spectrum offset, and only area random effects are estimated
$$
\text{logit} \left(\alpha_{x, s, a} \right) = u_x^\alpha + \eta_{R_x, s, a}^\alpha \\
$$
* If survey data on recent HIV infection are not included in the model, then $\beta_0^\lambda = \beta_S^{\lambda, s = \text{M}} = u_x^\lambda = 0$. The sex ratio for HIV incidence is determined by the sex incidence rate ratio from Spectrum in the same years and the incidence rate in all districts is modelled assuming the same average HIV transmission rate for untreated adults, but varies according to district estimates of HIV prevalence and ART coverage.

# \textsc{C++} template

```{cpp, echo=TRUE, eval=FALSE, output.var="ex", code=readLines("naomi.cpp")}

```

### 2022-12-21 src/docs_paper/paper.Rmd
A few appendix fixes and extra into main text

Let $i$ index the coarsest space-age-sex division included in the model.
Let $N_i$ be the population size and consider the following indicators: HIV prevalence $\rho_i$, annual HIV incidence rate $\lambda_i$ and ART coverage $\alpha_i$.

@@

To write.

>>>>

Let $i$ index the coarsest district-age-sex division included in the model.
Let $N_i$ be the population size and consider the following three indicators: HIV prevalence $\rho_i \in [0, 1]$, ART coverage $\alpha_i \in [0, 1]$, and annual HIV incidence rate $\lambda_i > 0$.

We specify independent mixed effects models for HIV prevalence and ART coverage on the logit scale such that
\begin{align}
\text{logit}(\rho_i) &= \eta^\rho_i, \\
\text{logit}(\alpha_i) &= \eta^\alpha_i,
\end{align}
for certain choice of linear predictors $\eta^\rho_i$ and $\eta^\alpha_i$.
For the HIV incidence rate we use a mixed effects model on the log scale where the linear predictor depends on $\{\rho_i, \alpha_i\}_{i \in \mathcal{I}}$
$$
\log(\lambda_i) = \eta^\rho_i(\{\rho_i, \alpha_i\}_{i \in \mathcal{I}}).
$$

@@

Consider a household survey occurring at $T_1$.
Let $\mathcal{I}$ be a set of $i$ for which observations are reported.
For each $\mathcal{I}$, we calculate the weighted observations $\hat \theta_\mathcal{I}$ for $\theta \in \{\rho, \alpha, \kappa\}$ with respective Kish effective sample sizes $M^{\hat \theta}_\mathcal{I}$, and observed number of cases
$$
Y^{\hat \theta}_{\mathcal{I}} = M^{\hat \theta}_{\mathcal{I}} \cdot \hat \theta_{\mathcal{I}}.
$$
